---
title: "MCMC"
output: rmarkdown::html_vignette
author: "Ray, Isaac"
date: "December 2021"
vignette: >
  %\VignetteIndexEntry{MCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(BASTION)
library(ggplot2)
library(ggraph)
library(igraph)
```

The purpose of this vignette is to demonstrate how the core functions of the `BASTION` package work individually, and then give a simple example of how they can be combined to predict responses at new data points.

# The Data & Motivation

We'll start by posing a situation. Suppose we are observing spatial points around a black hole, but have no data from within the black hole (our surveyors got too close). Below is a plot of all the locations at which we gathered data (stored in `data_hole`):

```{r data-gen, echo = FALSE}
set.seed(12345)
x = runif(1000, min = -10, max = 10)
y = runif(1000, min = -10, max = 10)
data_full = data.frame(x, y)
data_hole = data_full[x^2 + y^2 >= 16, ]
```

```{r data-plot, fig.width = 6, fig.height = 6}
ggplot(data_hole) + geom_point(aes(x, y))
```

At each data point, our space surveyors observed the value of the function we are interested in (stored in `response`). Below is a plot of the response data we observed at each point:


```{r data-response, echo = FALSE}
response = 1.2^(data_hole$x) + 
  (data_hole$y)^2 - 
  5*(data_hole$x > data_hole$y) + 
  4*data_hole$x*data_hole$y

```

```{r response-plot, fig.width = 6, fig.height = 6}
ggplot(data_hole) + 
  geom_point(aes(x, y, color = response)) +
  scale_color_viridis()
```

Unfortunately, all of our space surveyors have been sucked into the black hole. But we still need a way to predict the value of our function of interest at new data points. Before discussing how we will achieve this, let's discuss the core functions of `BASTION`.

# The constructGraph function

Before we can do any graph based procedures on our data, we need to have a graph! The `BASTION` package provides the function `constructGraph` to do this for us. `constructGraph` takes in two dimensional numeric data (in our case, `data_hole`) and a parameter `k` and performs the K-Nearest Neighbors algorithm to generate a weighted `igraph` graph. Below is a plot of what our graph looks like:

```{r make-graph, fig.width = 6, fig.height = 6}
hole_graph = constructGraph(data_hole, 5)
ggraph(hole_graph, layout = data_hole) + 
  geom_edge_link0(edge_alpha = 0.1) + 
  geom_node_point(aes(color = response)) +
  scale_color_viridis()
```

An important consideration is that the K-Nearest Neighbors algorithm is not guaranteed to generate a connected graph, especially in situations where there is not much data and the chosen `k` parameter is small. Other methods of constructing a graph, such as Constrained Delaunay triangulation, may be a better choice in such instances. The other functions of the `BASTION` package do not require that the original graph is made using the `constructGraph` function, a weighted `igraph` graph generated by other means will still work so long as it is connected and points share an edge with other points that are 'nearby' (we will be using Euclidean distance for simplicity).

# Initializing Clusters

Now that we have a graph, let's separate it into clusters. We are working on the principle that points which are 'close' together in space likely also have response values which are 'close' together. However, as we will see later, this assumption need not be true once we start performing different graph operations. With this in mind, the `constructClusters` function will form a partition of our graph into `nclust` different clusters (10 is chosen arbitrarily)  without needing to consider our `response` data.

The way that `constructClusters` creates the partition is by first using Prim's algorithm to find a minimum spanning tree on the input graph `hole_graph` using its edge weights (which was generated by `constructGraph`). By property of spanning trees, by removing `nclust - 1` edges from this spanning tree, we are left with `nclust` disconnected trees which together form a spanning forest. We then assign all the vertices in the same disconnected tree to the same cluster. The final spanning forest and cluster membership vector are both available in the list output by `constructClusters`.

Below is a demonstration of `constructClusters` on our `hole_graph` and a plot of which vertices were assigned to each cluster.

```{r make-cluster, fig.width = 6, fig.height = 6}
clustered_output = constructClusters(hole_graph, nclust = 10)
ggraph(clustered_output$spanning_forest, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(clustered_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

# Graph Operations

Now that we have a graph with clusters, we can demonstrate each of the four graph operations which are the core of the `BASTION` package.

## Birth Move

First, we will consider a cluster birth move. The `graphBirth` function, when provided with a clustered spanning forest graph and corresponding membership (`spanning_forest` and `membership` from the `constructClusters` output), will pick a random edge from the cluster indexed by `clust` and delete it. By deleting this edge, it splits the tree and creates an additional cluster and we say that a cluster is 'born'.

Note that `clust` is an optional parameter; cluster 1 was chosen for demonstration since in this example it happened to be the largest cluster. In practice `clust` is generally not specified, in which case a cluster is chosen at random from the existing clusters with at least one edge.


```{r cluster-birth, fig.width = 6, fig.height = 6}
birth_output = graphBirth(clustered_output$spanning_forest, clustered_output$membership, clust = 1)
ggraph(birth_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(birth_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

As we can see from the plot above, when compared with the original spanning forest graph, the original cluster 1 was split into the above clusters 1 and 3.

## Death Move

Next, we will consider the opposite of a cluster birth move: a cluster death move. The `graphDeath` function, when provided with a clustered spanning forest graph, corresponding membership, and the *original full graph* (`hole_graph` in this example), will randomly select an edge from the original graph which connects two clusters and add it back to the graph. In doing so, the two clusters that are connected become a single cluster and we say that a cluster has 'died'.

```{r cluster-death, fig.width = 6, fig.height = 6}
death_output = graphDeath(birth_output$graph, birth_output$membership, hole_graph)
ggraph(death_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(death_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

As we can see from the plot above, when compared with the plot from the birth move, the cluster which was born was subsequently killed (such is life for a cluster) and we are back to the original spanning forest we started with.

## Change Move

```{r cluster-change, fig.width = 6, fig.height = 6}
change_output = graphChange(death_output$graph, death_output$membership, hole_graph)
ggraph(change_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(change_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

Subtle change in groups 3, 9, 10

## Hyper Move

```{r cluster-hyper, fig.width = 6, fig.height = 6}
hyper_output = graphHyper(hole_graph, change_output$membership)
ggraph(hyper_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(hyper_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

# Simulation Study

Let's put it all together in a very simple estimation procedure. Suppose we want to try and estimate the response function as piecewise constant over the domain. Our estimate for the piecewise constant of each cluster will be the mean of the response at the cluster's member vertices, and the per-cluster error will be the sum of squared difference between each response and our estimate. We'll assess the total error of our graph by summing the error of each cluster, and penalize the number of clusters by adding a scaled exponential function of the number of clusters to the total error.

```{r estimation-functions}
estimation_procedure = function(membership, response) {
  k = length(unique(membership))
  estimates = rep(0, k)
  errors = rep(0, k)
  point_estimates = rep(0, length(membership))
  for(i in 1:k) {
    estimates[i] = mean(response[membership == i])
    errors[i] = sum((response[membership == i] - estimates[i])^2)
    point_estimates[membership == i] = estimates[i]
  }
  total_penalty = sum(errors) + (exp(k)/length(membership))
  return(list(total_penalty = total_penalty, 
              errors = errors, 
              estimates = estimates, 
              point_estimates = point_estimates))
}
```

Now, we'll repeatedly execute the 4 different graph operations on our original ```hole_graph``` with different probabilities. We can arbitrarily choose these probabilities; let's choose a 2/5 chance for a birth move, 2/5 for a death move, and 1/10 each for a hyper move or change move. In every iteration, we will select a move with the probabilities mentioned, evaluate if the error increases or decreases, and accept the new graph if its error decreased. Note that since edges play no role in our procedure, a hyper move will always be accepted as it does not affect the total penalty. 

```{r big-loop}
iterations = 2000
current_graph = clustered_output$spanning_forest
current_membership = clustered_output$membership
current_loss = estimation_procedure(current_membership, response)$total_penalty
for(i in 1:iterations) {
  move = sample(1:4, 1, prob = c(0.4, 0.4, 0.1, 0.1))
  if(move == 1) {
    proposed_output = graphBirth(current_graph, current_membership)
    proposed_estimation = estimation_procedure(proposed_output$membership, response)
    if(proposed_estimation$total_penalty <= current_loss) {
      current_graph = proposed_output$graph
      current_membership = proposed_output$membership
      current_loss = proposed_estimation$total_penalty
    }
  } else if(move == 2) {
    proposed_output = graphDeath(current_graph, current_membership, hole_graph)
    proposed_estimation = estimation_procedure(proposed_output$membership, response)
    if(proposed_estimation$total_penalty <= current_loss) {
      current_graph = proposed_output$graph
      current_membership = proposed_output$membership
      current_loss = proposed_estimation$total_penalty
    }
  } else if(move == 3) {
    proposed_output = graphChange(current_graph, current_membership, hole_graph)
    proposed_estimation = estimation_procedure(proposed_output$membership, response)
    if(proposed_estimation$total_penalty <= current_loss) {
      current_graph = proposed_output$graph
      current_membership = proposed_output$membership
      current_loss = proposed_estimation$total_penalty
    }
  } else if(move == 4) {
    proposed_output = graphHyper(hole_graph, current_membership)
    current_graph = proposed_output$graph
    current_membership = proposed_output$membership
  }
}
final_graph = current_graph
final_membership = current_membership
final_estimation_output = estimation_procedure(current_membership, response)
final_loss = final_estimation_output$total_penalty
final_errors = final_estimation_output$errors
```

Let's see how our estimation did. As a reminder, here is our original graph:
```{r original-graph, fig.width = 6, fig.height = 6}
ggraph(hole_graph, layout = data_hole) + 
  geom_edge_link0(edge_alpha = 0.1) + 
  geom_node_point(aes(color = response)) +
  scale_color_viridis() +
  labs(color = "Response")
```

And here are our final estimates:
```{r estimate-graph, fig.width = 7, fig.height = 6}
final_point_estimates = final_estimation_output$point_estimates
ggraph(final_graph, layout = data_hole) + 
  geom_edge_link0(edge_alpha = 0.1) + 
  geom_node_point(aes(color = final_point_estimates)) +
  scale_color_viridis() +
  labs(color = "Response Estimate")
```

## Predicting on Test Data

```{r prediction}
test_x = runif(500, min = -10, max = 10)
test_y = runif(500, min = -10, max = 10)
test_data_full = data.frame(x = test_x, y =  test_y)
test_data_hole = test_data_full[test_x^2 + test_y^2 >= 16, ]
test_response = 1.2^(test_data_hole$x) + 
  (test_data_hole$y)^2 - 
  5*(test_data_hole$x > test_data_hole$y) + 
  4*test_data_hole$x*test_data_hole$y

closest_point = function(new_point, existing_points) {
  return(unname(which.min(sqrt(colSums((new_point - t(existing_points))^2)))))
}

test_membership = rep(0, nrow(test_data_hole))
for(i in 1:nrow(test_data_hole)) {
  test_membership[i] = final_membership[closest_point(as.numeric(test_data_hole[i,]), data_hole)]
}

test_estimation_output = estimation_procedure(test_membership, test_response)
test_pred_response = test_estimation_output$point_estimates
```


```{r true-test-graph, fig.width = 7, fig.height = 6}
ggplot(test_data_hole) + 
  geom_point(aes(x, y, color = test_response)) +
  scale_color_viridis(limits = c(-300, 500)) +
  labs(color = "Response")
```

```{r pred-test-graph, fig.width = 7, fig.height = 6}
ggplot(test_data_hole) + 
  geom_point(aes(x, y, color = test_pred_response)) +
  scale_color_viridis(limits = c(-300, 500)) +
  labs(color = "Predicted Response")
```

```{r residual, fig.width = 7, fig.height = 6}
test_error = test_response - test_pred_response
ggplot() + 
  geom_histogram(aes(x = test_error), bins = 20) +
  labs(x = "Prediction Error")
```




