---
title: "MCMC"
output: rmarkdown::html_vignette
author: "Ray, Isaac"
date: "December 2021"
vignette: >
  %\VignetteIndexEntry{MCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BASTION)
library(ggplot2)
library(ggraph)
library(igraph)
```

```{r data-gen, fig.width = 6, fig.height = 6}
set.seed(12345)
x = runif(500, min = -10, max = 10)
y = runif(500, min = -10, max = 10)
data_full = data.frame(x, y)
data_hole = data_full[x^2 + y^2 >= 16, ]
ggplot(data_hole) + geom_point(aes(x, y))
```

```{r data-response, fig.width = 6, fig.height = 6}
response = 1.2^(data_hole$x) + (data_hole$y)^2 - 5*(data_hole$x > data_hole$y) + 4*data_hole$x*data_hole$y
ggplot(data_hole) + 
  geom_point(aes(x, y, color = response)) +
  scale_color_viridis()
```

```{r make-graph, fig.width = 6, fig.height = 6}
hole_graph = constructGraph(data_hole, 5)
ggraph(hole_graph, layout = data_hole) + 
  geom_edge_link0(edge_alpha = 0.1) + 
  geom_node_point(aes(color = response)) +
  scale_color_viridis()
```

```{r make-cluster, fig.width = 6, fig.height = 6}
clustered_output = constructClusters(hole_graph, nclust = 10)
ggraph(clustered_output$spanning_forest, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(clustered_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

```{r cluster-birth, fig.width = 6, fig.height = 6}
birth_output = graphBirth(clustered_output$spanning_forest, clustered_output$membership, clust = 7)
ggraph(birth_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(birth_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

```{r cluster-death, fig.width = 6, fig.height = 6}
death_output = graphDeath(birth_output$graph, birth_output$membership, hole_graph)
ggraph(death_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(death_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

```{r cluster-change, fig.width = 6, fig.height = 6}
change_output = graphChange(death_output$graph, death_output$membership, hole_graph)
ggraph(change_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(change_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

Subtle change in groups 3, 9, 10

```{r cluster-hyper, fig.width = 6, fig.height = 6}
hyper_output = graphHyper(hole_graph, change_output$membership)
ggraph(hyper_output$graph, layout = data_hole) +
  geom_edge_link0(edge_alpha = 0.1) +
  geom_node_point(aes(colour = as.factor(hyper_output$membership))) +
  scale_colour_brewer(palette = "Paired") + 
  labs(colour = "Cluster")
```

Let's put it all together in a very simple estimation procedure. Suppose we want to try and estimate the response function as piecewise constant over the domain. Our estimate for the piecewise constant of each cluster will be the mean of the response at the cluster's member vertices, and the per-cluster error will be the sum of squared difference between each response and our estimate. We'll assess the total error of our graph by summing the error of each cluster, and penalize the number of clusters by adding a scaled exponential function of the number of clusters to the total error.

```{r estimation-functions}
estimation_procedure = function(membership, response) {
  k = length(unique(membership))
  estimates = rep(0, k)
  errors = rep(0, k)
  point_estimates = rep(0, length(membership))
  for(i in 1:k) {
    estimates[i] = mean(response[membership == i])
    errors[i] = sum((response[membership == i] - estimates[i])^2)
    point_estimates[membership == i] = estimates[i]
  }
  total_penalty = sum(errors) + (exp(k)/length(membership))
  return(list(total_penalty = total_penalty, errors = errors, estimates = estimates, point_estimates = point_estimates))
}
```

Now, we'll repeatedly execute the 4 different graph operations on our original ```hole_graph``` with different probabilities. We can arbitrarily choose these probabilities; let's choose a 2/5 chance for a birth move, 2/5 for a death move, and 1/10 each for a hyper move or change move. In every iteration, we will select a move with the probabilities mentioned, evaluate if the error increases or decreases, and accept the new graph if its error decreased. Note that since edges play no role in our procedure, a hyper move will always be accepted as it does not affect the total penalty. 

```{r big-loop}
iterations = 500
current_graph = clustered_output$spanning_forest
current_membership = clustered_output$membership
current_loss = estimation_procedure(current_membership, response)$total_penalty
for(i in 1:500) {
  move = sample(1:4, 1, prob = c(0.4, 0.4, 0.1, 0.1))
  if(move == 1) {
    proposed_output = graphBirth(current_graph, current_membership)
    proposed_estimation = estimation_procedure(proposed_output$membership, response)
    if(proposed_estimation$total_penalty <= current_loss) {
      current_graph = proposed_output$graph
      current_membership = proposed_output$membership
      current_loss = proposed_estimation$total_penalty
    }
  } else if(move == 2) {
    proposed_output = graphDeath(current_graph, current_membership, hole_graph)
    proposed_estimation = estimation_procedure(proposed_output$membership, response)
    if(proposed_estimation$total_penalty <= current_loss) {
      current_graph = proposed_output$graph
      current_membership = proposed_output$membership
      current_loss = proposed_estimation$total_penalty
    }
  } else if(move == 3) {
    proposed_output = graphChange(current_graph, current_membership, hole_graph)
    proposed_estimation = estimation_procedure(proposed_output$membership, response)
    if(proposed_estimation$total_penalty <= current_loss) {
      current_graph = proposed_output$graph
      current_membership = proposed_output$membership
      current_loss = proposed_estimation$total_penalty
    }
  } else if(move == 4) {
    proposed_output = graphHyper(hole_graph, current_membership)
    current_graph = proposed_output$graph
    current_membership = proposed_output$membership
  }
}
final_graph = current_graph
final_membership = current_membership
final_estimation_output = estimation_procedure(current_membership, response)
final_loss = final_estimation_output$total_penalty
final_errors = final_estimation_output$errors
```

Let's see how our estimation did. As a reminder, here is our original graph:
```{r original-graph, fig.width = 6, fig.height = 6}
ggraph(hole_graph, layout = data_hole) + 
  geom_edge_link0(edge_alpha = 0.1) + 
  geom_node_point(aes(color = response)) +
  scale_color_viridis() +
  labs(color = "Response")
```

And here are our final estimates:
```{r estimate-graph, fig.width = 7, fig.height = 6}
final_point_estimates = final_estimation_output$point_estimates
ggraph(final_graph, layout = data_hole) + 
  geom_edge_link0(edge_alpha = 0.1) + 
  geom_node_point(aes(color = final_point_estimates)) +
  scale_color_viridis() +
  labs(color = "Response Estimate")
```








