---
title: "Speed Comparisons"
output: rmarkdown::html_vignette
author: "Ray, Isaac"
date: "January 2022"
vignette: >
  %\VignetteIndexEntry{speed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r setup, include = FALSE}
library(BASTION)
library(ggplot2)
library(ggraph)
library(igraph)
library(microbenchmark)
library(Rfast)
library(xtable)
library(fields)
library(Matrix)
library(FNN)
library(mgcv)
library(fdaPDE)
```

In the effort to improve the computational efficiency of the BAST model, this vignette summarizes various implementations and their speed. 

First, a speed comparison on the remote sensed chlorophyll data in the Aral sea; the setup of this application is described in section 4.3 of the following paper:

[Luo, Z. T., Sang, H., & Mallick, B. (2021) BAST: Bayesian Additive Regression Spanning Trees for Complex Constrained Domain. Advances in Neural Information Processing Systems 34 (NeurIPS 2021)](https://proceedings.neurips.cc/paper/2021/hash/00b76fddeaaa7d8c2c43d504b2babd8a-Abstract.html)

The code for the paper's implementation can be found [in this GitHub repository](https://github.com/ztluostat/BAST). The majority of the code for this example is taken from this repository.


```{r ztluo-funcs, include=FALSE}
### Redefine functions related to FEM ###
### code adapted from fdaPDE ###

# function to returns the index of the triangle containing the point
# returns multiple triangles for points on edges if all_tri == TRUE
# code adapted from fdaPDE
R_insideIndex2 <- function (mesh, location, all_tri = FALSE) {
  #  insideIndex returns the index of the triangle containing the point
  # (X,Y) if such a triangle exists, and NaN otherwise.
  #  TRICOEF may have already been calculated for efficiency,
  #  but if the function is called with four arguments, it is calculated.
  eps = 2.2204e-016
  small = 10000 * eps
  
  nodes = mesh$nodes
  triangles = mesh$triangles
  X = location[1]
  Y = location[2]
  
  ntri   = dim(triangles)[[1]]
  indtri   = matrix(1:ntri, ncol = 1)
  
  #  compute coefficients for computing barycentric coordinates if needed
  
  tricoef = fdaPDE:::R_tricoefCal(mesh)
  
  #  compute barycentric coordinates
  r3 = X - nodes[triangles[, 3], 1]
  s3 = Y - nodes[triangles[, 3], 2]
  lam1 = (tricoef[, 4] * r3 - tricoef[, 2] * s3)
  lam2 = (-tricoef[, 3] * r3 + tricoef[, 1] * s3)
  lam3 = 1 - lam1 - lam2
  
  #  test these coordinates for a triple that are all between 0 and 1
  int  = (-small <= lam1 & lam1 <= 1 + small) &
    (-small <= lam2 & lam2 <= 1 + small) &
    (-small <= lam3 & lam3 <= 1 + small)
  
  #  return the index of this triple, or NaN if it doesn't exist
  indi = indtri[int]
  if (length(indi) < 1)
  {
    ind = NA
  } else if (all_tri) {
    ind = indi
  } else{
    ind = min(indi)
  }
  
  return(ind)
}
### Functions for fitting BAST -----

# function to get whether an edge is within a cluster or bewteen two clusters
getEdgeStatus <- function(membership, inc_mat) {
  membership_head = membership[inc_mat[, 1]]
  membership_tail = membership[inc_mat[, 2]]
  edge_status = rep('w', nrow(inc_mat))
  edge_status[membership_head != membership_tail] = 'b'
  return(edge_status)
}

# function to split an existing cluster given MST
# subgraphs: cluster_id -> subgraph
# cluster: vid -> cluster_id
splitCluster <- function(mstgraph, k, subgraphs, csize) { 
  clust_split = sample.int(k, 1, prob = csize - 1)
  mst_subgraph = subgraphs[[clust_split]]
  
  edge_cutted = E(mst_subgraph)[sample.int(csize[clust_split]-1, 1)]
  eid_cutted = edge_cutted$eid
  mst_subgraph = delete.edges(mst_subgraph, edge_cutted)
  connect_comp = components(mst_subgraph)
  idx_new = (connect_comp$membership == 2)
  vid_new = V(mst_subgraph)$vid[idx_new]
  vid_old = V(mst_subgraph)$vid[!idx_new]
  
  return(list(vid_old = vid_old, vid_new = vid_new, eid_cutted = eid_cutted,
              clust_old = clust_split, idx_new = idx_new))
}

# function to update if a split move is accepted
updateSplit <- function(split_res, subgraphs, k, csize, eid_btw_mst, cluster, edge_status,
                        adj_list, adj_edge_list) {
  clust_split = split_res$clust_old
  vid_old = split_res$vid_old; vid_new = split_res$vid_new
  
  subgraph_split = subgraphs[[clust_split]]
  idx_new = split_res$idx_new
  subgraphs[[clust_split]] = induced_subgraph(subgraph_split, !idx_new)  # subgraph of old cluster
  subgraphs[[k+1]] = induced_subgraph(subgraph_split, idx_new) # subgraph of new cluster
  
  csize[clust_split] = length(vid_old)
  csize[k+1] = length(vid_new)
  
  cluster[vid_new] = k + 1
  eid_btw_mst = c(eid_btw_mst, split_res$eid_cutted)
  
  # update edge status
  adj_vid_old = unlist(adj_list[vid_old])
  adj_eid_old = unlist(adj_edge_list[vid_old])
  clust_adj_old = cluster[adj_vid_old]
  idx_btw = which(clust_adj_old != clust_split)
  eid_btw = adj_eid_old[idx_btw]
  edge_status[eid_btw] = 'b'
  
  return(list(subgraphs = subgraphs, csize = csize, cluster = cluster,
              eid_btw_mst = eid_btw_mst, estatus = edge_status))
}

# function to merge two existing clusters
mergeCluster <- function(mstgraph, eid_btw_mst, subgraphs, csize, cluster, edge_list, 
                         change = F) {
  # edge for merging
  edge_merge = sample.int(length(eid_btw_mst), 1)
  # update cluster information
  # clusters of endpoints of edge_merge
  eid_merge = eid_btw_mst[edge_merge]
  clusters_merge = cluster[edge_list[eid_merge, ]]
  clusters_merge = sort(clusters_merge)
  c1 = clusters_merge[1]; c2 = clusters_merge[2] # note c1 < c2
  # merge c2 to c1
  
  # vid of vertices in c2
  vid_old = V(subgraphs[[c2]])$vid
  # vid in merged cluster
  vid_new = c(V(subgraphs[[c1]])$vid, vid_old)
  
  csize_new = NULL; subgraphs_new = NULL
  if(change) {
    subgraphs_new = subgraphs
    subgraphs_new[[c1]] = induced_subgraph(mstgraph, vid_new)
    subgraphs_new[[c2]] = NULL
    
    csize_new = csize
    csize_new[c1] = length(vid_new)
    csize_new = csize_new[-c2]
  }
  
  # now drop c2
  return(list(vid_old = vid_old, vid_new = vid_new, clust_old = c2, clust_new = c1,
              edge_merge = edge_merge, subgraphs = subgraphs_new, csize = csize_new))
}

# function to update if a merge move is accepted
updateMerge <- function(res_merge, subgraphs, csize, eid_btw_mst, cluster, edge_status,
                        adj_list, adj_edge_list, mstgraph) {
  clust_old = res_merge$clust_old; clust_new = res_merge$clust_new
  vid_old = V(subgraphs[[clust_old]])$vid
  vid_new = c(V(subgraphs[[clust_new]])$vid, vid_old)
  subgraphs[[clust_new]] = induced_subgraph(mstgraph, vid_new)
  subgraphs[[clust_old]] = NULL
  
  csize[clust_new] = length(vid_new)
  csize = csize[-clust_old]
  
  cluster[vid_old] = clust_new
  idx = which(cluster > clust_old)
  cluster[idx] = cluster[idx] - 1
  
  eid_btw_mst = eid_btw_mst[-res_merge$edge_merge]
  
  # update edge status
  adj_vid_old = unlist(adj_list[vid_old])
  adj_eid_old = unlist(adj_edge_list[vid_old])
  clust_adj_old = cluster[adj_vid_old]
  idx_within = which(clust_adj_old == clust_new)
  eid_within = adj_eid_old[idx_within]
  edge_status[eid_within] = 'w'
  
  return(list(subgraphs = subgraphs, csize = csize, cluster = cluster,
              eid_btw_mst = eid_btw_mst, estatus = edge_status))
}

# function to propose a new MST
proposeMST <- function(graph0, edge_status, subgraphs) {
  nedge = length(edge_status)
  nb = sum(edge_status == 'b')
  nw = nedge - nb
  weight = numeric(nedge)
  weight[edge_status == 'w'] = runif(nw)
  weight[edge_status == 'b'] = runif(nb, 10, 20)
  E(graph0)$weight = weight
  mstgraph = mst(graph0, weights = E(graph0)$weight)
  
  # update subgraphs
  subgraphs_new = lapply(subgraphs, function(g, mstgraph) {induced_subgraph(mstgraph, V(g)$vid)},
                         mstgraph)
  # update eid_btw_mst
  eid_btw_mst = E(mstgraph)$eid[E(mstgraph)$weight >= 10]
  
  return(list(mstgraph = mstgraph, subgraphs = subgraphs_new, eid_btw_mst = eid_btw_mst))
}

# function to get log likelihood ratio
evalLogLikeRatio <- function(move, e_m, vid_old, vid_new, sigmasq_y, sigmasq_mu) {
  sigma_ratio = sigmasq_y / sigmasq_mu
  csize_old = length(vid_old); csize_new = length(vid_new)
  sum_e_old = sum(e_m[vid_old]); sum_e_new = sum(e_m[vid_new])
  
  if(move == 'split') {
    logdetdiff = -0.5 * (log(csize_old+sigma_ratio) + log(csize_new+sigma_ratio) - 
                           log(csize_old+csize_new+sigma_ratio) - log(sigma_ratio))
    quaddiff = 0.5 * (sum_e_old^2/(csize_old+sigma_ratio) + sum_e_new^2/(csize_new+sigma_ratio) - 
                        (sum_e_old+sum_e_new)^2/(csize_old+csize_new+sigma_ratio)) / sigmasq_y
  }
  
  if(move == 'merge') {
    logdetdiff = -0.5 * (log(csize_new+sigma_ratio) - log(csize_old+sigma_ratio) - 
                           log(csize_new-csize_old+sigma_ratio) + log(sigma_ratio))
    quaddiff = 0.5 * (sum_e_new^2/(csize_new+sigma_ratio) - sum_e_old^2/(csize_old+sigma_ratio) - 
                        (sum_e_new-sum_e_old)^2/(csize_new-csize_old+sigma_ratio)) / sigmasq_y
  }
  return(logdetdiff + quaddiff)
}

# function to get log posterior density (up to a constant)
evalLogPost <- function(mu, mu_all, sigmasq_y, k, Y, hyper) {
  n = length(Y); M = length(k)
  sigmasq_mu = hyper[1]; lambda_s = hyper[2]; nu = hyper[3]; lambda_k = hyper[4]
  log_prior =  -(nu/2+1)*log(sigmasq_y) - nu*lambda_s/(2*sigmasq_y)
  log_prior = log_prior + sum(-lchoose(n-1, k-1) + k*log(lambda_k) - lfactorial(k))
  log_prior = log_prior - sum(unlist(mu)^2) / (2*sigmasq_mu)
  
  Y_hat = rowSums(mu_all)
  log_like = -n/2*log(sigmasq_y) - sum((Y - Y_hat)^2) / (2*sigmasq_y)
  return(log_like + log_prior)
}

# function to standardize Y
standardize <- function(x) {
  xmean = mean(x)
  x = x - xmean
  xscale = 2 * max(abs(x))
  x = x / xscale
  param = c('mean' = xmean, 'scale' = xscale)
  return(list(x = x, std_par = param))
}

# function to unstandardize Y
unstandardize <- function(x, std_par, nomean = F, s2 = F) {
  if(s2) {
    x = x * std_par['scale'] ^ 2
  } else {
    x = x * std_par['scale']
  }
  if(!nomean) x = x + std_par['mean']
  return(x)
}
fitBAST <- function(Y, graph0, init_val, hyperpar, MCMC, BURNIN, THIN, seed = 1234) {
  set.seed(seed)
  
  n = vcount(graph0)
  # hyper-parameter
  M = hyperpar['M']  # number of trees
  sigmasq_mu = hyperpar['sigmasq_mu']
  lambda_s = hyperpar['lambda_s']
  nu = hyperpar['nu']
  lambda_k = hyperpar['lambda_k']
  k_max = hyperpar['k_max']
  hyper = c(sigmasq_mu, lambda_s, nu, lambda_k)
  
  if('name' %in% names(vertex_attr(graph0))) {
    graph0 = delete_vertex_attr(graph0, 'name')
  }
  inc_mat = get.edgelist(graph0, names = F)
  adj_list = lapply(as_adj_list(graph0), FUN = function(x) {x$vid})
  adj_edge_list = lapply(as_adj_edge_list(graph0), FUN = function(x) {x$eid})
  
  
  # initial values
  mstgraph_lst = init_val[['trees']]
  mu = init_val[['mu']]
  cluster = init_val[['cluster']]  # n*M matrix
  sigmasq_y = init_val[['sigmasq_y']]
  k = as.numeric(apply(cluster, 2, max))  # number of clusters
  
  csize = list() # cluster size
  subgraphs = list()
  eid_btw_mst = list()
  g = matrix(0, nrow = n, ncol = M)  # n*M matrix of fitted mu's
  for(m in 1:M) {
    cluster_m = cluster[, m]
    g[, m] = mu[[m]][cluster_m]
    
    csize[[m]] = Rfast::Table(cluster_m)
    
    mstgraph_m = mstgraph_lst[[m]]
    clust_vid_m = split(1:n, cluster_m)
    subgraphs[[m]] = lapply(clust_vid_m, function(vids, mstgraph) {
      induced_subgraph(mstgraph, vids)
    }, mstgraph_m)
    
    inc_mat_mst = get.edgelist(mstgraph_m, names = F)
    # idx for bewteen edges
    c1_m = cluster_m[inc_mat_mst[, 1]]; c2_m = cluster_m[inc_mat_mst[, 2]]
    idx_btw = which(c1_m != c2_m)
    eid_btw_mst[[m]] = (E(mstgraph_m)$eid)[idx_btw]
  }
  
  # whether an edge in graph0 is within a cluster or bewteen two clusters
  # n*M matrix
  edge_status = apply(cluster, 2, FUN = getEdgeStatus, inc_mat)
  
  ################# MCMC ####################
  
  ## MCMC results
  mu_out = list()
  sigmasq_y_out = numeric((MCMC-BURNIN)/THIN)
  cluster_out = array(0, dim = c((MCMC-BURNIN)/THIN, n, M))
  tree_out = list()
  log_post_out = numeric((MCMC-BURNIN)/THIN)
  
  ## MCMC iteration
  for(iter in 1:MCMC) {
    for(m in 1:M) {
      k_m = k[m]
      mstgraph_m = mstgraph_lst[[m]]
      edge_status_m = edge_status[, m]
      cluster_m = cluster[, m]
      csize_m = csize[[m]]
      subgraphs_m = subgraphs[[m]]
      eid_btw_mst_m = eid_btw_mst[[m]]
      
      e_m = Y - Rfast::rowsums(g[, -m])
      
      if(k_m == 1) {rb = 0.9; rd = 0; rc = 0; rhy = 0.1
      } else if(k_m == min(k_max, n)) {rb = 0; rd = 0.6; rc = 0.3; rhy = 0.1
      } else {rb = 0.3; rd = 0.3; rc = 0.3; rhy = 0.1}
      move = sample(4, 1, prob = c(rb, rd, rc, rhy))
      
      if(move == 1) { ## Birth move
        # split an existing cluster
        split_res = splitCluster(mstgraph_m, k_m, subgraphs_m, csize_m)
        vid_new = split_res$vid_new; vid_old = split_res$vid_old
        
        # compute log-prior ratio
        log_A = log(lambda_k) - log(k_m + 1)
        # compute log-proposal ratio
        if(k_m == min(k_max, n)-1) {
          rd_new = 0.6
        } else {rd_new = 0.3}
        log_P = log(rd_new) - log(rb)
        # compute log-likelihood ratio
        log_L = evalLogLikeRatio('split', e_m, vid_old, vid_new, sigmasq_y, sigmasq_mu)
        
        #acceptance probability
        acc_prob = min(0, log_A + log_P + log_L)
        acc_prob = exp(acc_prob)
        if(runif(1) < acc_prob){
          # accept
          update_res = updateSplit(split_res, subgraphs_m, k_m, csize_m, eid_btw_mst_m, 
                                   cluster_m, edge_status_m, adj_list, adj_edge_list)
          subgraphs[[m]] = update_res$subgraphs
          csize[[m]] = update_res$csize
          eid_btw_mst[[m]] = update_res$eid_btw_mst
          cluster[, m] = update_res$cluster
          k[m] = k[m] + 1
          edge_status[, m] = update_res$estatus
        }
      }
      
      if(move == 2) { ## Death move
        # merge two existing clusters (c1, c2) -> c2
        merge_res = mergeCluster(mstgraph_m, eid_btw_mst_m, subgraphs_m, csize_m, 
                                 cluster_m, inc_mat)
        vid_old = merge_res$vid_old; vid_new = merge_res$vid_new
        
        # compute log-prior ratio
        log_A = -log(lambda_k) + log(k_m)
        # # compute log-proposal ratio
        if(k_m == 2) {rb_new = 0.9
        }else {rb_new = 0.3}
        log_P = -(log(rd) - log(rb_new))
        # compute log-likelihood ratio
        log_L = evalLogLikeRatio('merge', e_m, vid_old, vid_new, sigmasq_y, sigmasq_mu)
        
        # acceptance probability
        acc_prob = min(0, log_A + log_P + log_L)
        acc_prob = exp(acc_prob)
        if(runif(1) < acc_prob){
          # accept
          update_res = updateMerge(merge_res, subgraphs_m, csize_m, eid_btw_mst_m, 
                                   cluster_m, edge_status_m, adj_list, adj_edge_list, mstgraph_m)
          subgraphs[[m]] = update_res$subgraphs
          csize[[m]] = update_res$csize
          eid_btw_mst[[m]] = update_res$eid_btw_mst
          cluster[, m] = update_res$cluster
          k[m] = k[m] - 1
          edge_status[, m] = update_res$estatus
        }
      }
      
      if(move == 3) { ## change move
        # first perform death move: (c1, c2) -> c2
        merge_res = mergeCluster(mstgraph_m, eid_btw_mst_m, subgraphs_m, csize_m, 
                                 cluster_m, inc_mat, change = T)
        # then perform birth move
        split_res = splitCluster(mstgraph_m, k_m-1, merge_res$subgraphs, merge_res$csize)
        
        # compute log-likelihood ratio
        log_L = evalLogLikeRatio('merge', e_m, merge_res$vid_old, merge_res$vid_new, sigmasq_y, sigmasq_mu) + 
          evalLogLikeRatio('split', e_m, split_res$vid_old, split_res$vid_new, sigmasq_y, sigmasq_mu)
        
        # acceptance probability
        acc_prob = min(0, log_L)
        acc_prob = exp(acc_prob)
        if(runif(1) < acc_prob){
          # accept
          update_res = updateMerge(merge_res, subgraphs_m, csize_m, eid_btw_mst_m, 
                                   cluster_m, edge_status_m, adj_list, adj_edge_list, mstgraph_m)
          update_res = updateSplit(split_res, update_res$subgraphs, k_m-1, update_res$csize, 
                                   update_res$eid_btw_mst, update_res$cluster, 
                                   update_res$estatus, adj_list, adj_edge_list)
          
          subgraphs[[m]] = update_res$subgraphs
          csize[[m]] = update_res$csize
          eid_btw_mst[[m]] = update_res$eid_btw_mst
          cluster[, m] = update_res$cluster
          edge_status[, m] = update_res$estatus
        }
      }
      
      if(move == 4) {
        # update MST
        mstgraph_m = proposeMST(graph0, edge_status[, m], subgraphs_m)
        mstgraph_lst[[m]] = mstgraph_m$mstgraph
        
        # update eid_btw_mst
        eid_btw_mst[[m]] = mstgraph_m$eid_btw_mst
        # update subgraphs
        subgraphs[[m]] = mstgraph_m$subgraphs
      }
      
      # update mu_m
      k_m = k[m]; cluster_m = cluster[, m]; csize_m = csize[[m]]
      Qinv_diag = 1 / (csize_m/sigmasq_y + 1/sigmasq_mu)
      b = Qinv_diag * Rfast::group.sum(e_m, cluster_m) / sigmasq_y
      mu[[m]] = rnorm(k_m, b, sqrt(Qinv_diag))
      
      g[, m] = mu[[m]][cluster_m]
    }
    
    # update sigmasq_y
    Y_hat = g[, M] + Y - e_m
    rate = 0.5*(nu*lambda_s + sum((Y - Y_hat)^2))
    sigmasq_y = 1/rgamma(1, shape = (n+nu)/2, rate = rate)
    
    
    ## save result
    if(iter > BURNIN & (iter - BURNIN) %% THIN == 0) {
      mu_out[[(iter-BURNIN)/THIN]] = mu
      sigmasq_y_out[(iter-BURNIN)/THIN] = sigmasq_y
      #tree_out[[(iter-BURNIN)/THIN]] = mstgraph_lst
      cluster_out[(iter-BURNIN)/THIN, , ] = cluster
      
      log_post_out[(iter-BURNIN)/THIN] = evalLogPost(mu, g, sigmasq_y, k, Y, hyper)
    }
    
    if(iter %% 100 == 0)
      cat('Iteration', iter, 'done\n')
  }
  
  mode(cluster_out) = 'integer'  # to save memory
  return(list('mu_out' = mu_out,
              'sigmasq_y_out' = sigmasq_y_out,
              'cluster_out' = cluster_out, 'log_post_out' = log_post_out))
}
### Functions for 2-d constrained domains -----

# function to create a 2d mesh
gen2dMesh <- function(coords, bnd, ...) {
  # note the first and last boundary nodes are the same
  n = nrow(coords); n_bnd = length(bnd$x) - 1
  coords_all = rbind(coords, cbind(bnd$x, bnd$y)[1:n_bnd, ])
  
  # get boundary segments
  segments = cbind( (n+1):(n+n_bnd), c((n+2):(n+n_bnd), n+1) )
  
  mesh = create.mesh.2D(coords_all, segments = segments, ...)
  mesh$n_int = n  # number of interior nodes
  # edges that connect boundary nodes
  mesh$bnd_edges = apply(mesh$edges, 1, FUN = function(x) any(x > n))
  return(mesh)
}

# function to obtain a constrained Delaunay triangulation graph from a mesh
constrainedDentri <- function(n, mesh, threshold = 1000) {
  coords = mesh$nodes[1:n, ]
  
  # drop edges that connect boundary nodes
  rid_drop = mesh$bnd_edges
  edge_list = mesh$edges[!rid_drop, ]
  
  # compute edge length
  distance = sqrt( rowSums((coords[edge_list[, 1], ] - coords[edge_list[, 2], ]) ^ 2) )
  
  rid_drop = distance > threshold
  edge_list = edge_list[!rid_drop, ]
  distance = distance[!rid_drop]
  
  graph0 = graph_from_edgelist(edge_list, directed = F)
  E(graph0)$weight = distance
  return(graph0)
}

# function to generate equally spaced points along a given line
refineLine <- function(start, end, n) {
  grids_x = seq(start[1], end[1], length.out = n)
  grids_y = seq(start[2], end[2], length.out = n)
  grids = cbind(grids_x, grids_y)
  return(grids)
}

### Constrained KNN -----

# function to get a KNN graph given a distance matrix
KNNGraph <- function(dist_mat, k_nn = 5, cross_dist = F, return_graph = T) {
  
  n1 = nrow(dist_mat); n2 = ncol(dist_mat)
  if(cross_dist) {
    # dist_mat is cross distance matrix
    adj_list = apply(dist_mat, 1, function(x) order(x)[1:k_nn])
  } else {
    adj_list = apply(dist_mat, 1, function(x) order(x)[2:(k_nn+1)])
  }
  adj_list = t(adj_list)
  
  if(return_graph & n1 == n2) {
    require(igraph)
    require(Matrix)
    
    i_all = c(); j_all = c(); x_all = c()
    for(i in 1:n1) {
      for(cidx in 1:k_nn) {
        i_all = c(i_all, i)
        j_all = c(j_all, adj_list[i, cidx])
        x_all = c(x_all, dist_mat[i, adj_list[i, cidx] ])
      }
    }
    adj_mat = sparseMatrix(i = i_all, j = j_all, x = x_all, dims = c(n1, n2))
    # get knn graph
    knngraph = graph_from_adjacency_matrix(adj_mat, mode = 'directed', weighted = T)
    knngraph = as.undirected(knngraph, mode = 'collapse', edge.attr.comb = 'first')
    return(knngraph)
  } else {
    return(lapply(1:n1, function(i) adj_list[i, ]))
  }
}


### Plotting functions -----

# function to plot complex domain
geom_boundary <- function(bnd, ...) {
  n = length(bnd$x)
  segments = cbind(bnd$x[-n], bnd$y[-n], bnd$x[-1], bnd$y[-1])
  segments = data.frame(segments)
  names(segments) = c('x1', 'y1', 'x2', 'y2')
  return(geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = segments, ...))
}

# plotting functions from SCC
plotGraph <- function(coords, graph, title = NULL){
  require(ggplot2)
  edgelist = get.edgelist(graph) 
  edgedata = data.frame(coords[edgelist[,1 ], ], coords[edgelist[, 2], ])
  colnames(edgedata) = c("X1", "Y1", "X2", "Y2")
  
  ggplot() + geom_segment(aes(x=X1, y=Y1, xend = X2, yend = Y2), data = edgedata, size = 0.5, colour = "grey") +
    labs(title = title, x = "lon", y = "lat")+
    theme(plot.title = element_text(hjust = 0.5))
}

# function to plot spatial field
plotField <- function(coords, Data, col_lim = NULL, legend_name = NULL, title = NULL, colors = rainbow(10)){
  if(missing(col_lim)) {col_lim = range(Data)}
  ggplot() + 
    geom_tile(data = data.frame(coords), aes(lon, lat, fill = Data)) +
    scale_fill_gradientn(colours = colors, limits = col_lim, name = legend_name, na.value = 'white') +
    ggtitle(title) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "black"),
          # legend.title=element_blank(), 
          plot.title = element_text(hjust = 0.5, size = 10),
          legend.text = element_text(size = 9, hjust = 0, margin = margin(l = 3)))
}
```

First, we'll read in the processed data (information on this can be found in the AralSea.R file at the GitHub repository above)

```{r get-aral-data}
aralDataUrl = "https://github.com/ztluostat/BAST/raw/main/aral_data.RData"
load(url(aralDataUrl))
n = length(Y)
```

Here is a plot of what the data looks like:

```{r aral-plot, fig.width = 6, fig.height = 6}
ggplot() + 
  geom_boundary(bnd) +
  geom_point(aes(x = lon, y = lat, col = Y), data = as.data.frame(coords)) +
  scale_color_gradientn(colours = rainbow(5), name = 'Y') +
  labs(x = 'Scaled Lon.', y = 'Scaled Lat.') + 
  ggtitle('Observed Data')
```

The BASTION package does not include a function for constructing a constrained Delauney triangulation graph, so the graph will be constructed using the `constrainedDentri` function from the paper repository and used for all implementations.

```{r make-aral-graph}
aral_graph = constrainedDentri(n, gen2dMesh(coords, bnd)) # Get full graph
E(aral_graph)$eid = c(1:ecount(aral_graph)) # Assign edge ids
V(aral_graph)$vid = c(1:vcount(aral_graph)) # Assign vertex ids
# plot spatial graph
plotGraph(coords, aral_graph) + 
  geom_boundary(bnd) +
  labs(x = 'Scaled Lon.', y = 'Scaled Lat.') + 
  ggtitle('Spatial Graph')
```

The following is the implementation from the original paper. In the interest of timing, the iteration numbers have been cut down and such changes are commented below.

```{r ztluo-implement}
mstgraph = mst(aral_graph)  # initial spanning tree
graph0 = delete_edge_attr(aral_graph, 'weight')
mstgraph0 = delete_edge_attr(mstgraph, 'weight')

M = 30      # number of weak learners
k_max = 5   # maximum number of clusters per weak learner
mu = list() # initial values of mu
mstgraph_lst = list()  # initial spanning trees
cluster = matrix(1, nrow = n, ncol = M)  # initial cluster memberships
for(m in 1:M) {
  mu[[m]] = c(0)
  mstgraph_lst[[m]] = mstgraph0
}

init_val = list()
init_val[['trees']] = mstgraph_lst
init_val[['mu']] = mu
init_val[['cluster']] = cluster
init_val[['sigmasq_y']] = 1

# standardize Y
std_res = standardize(Y)
Y_std = std_res$x
std_par = std_res$std_par

# find lambda_s
nu = 3; q = 0.9
quant = qchisq(1-q, nu)
lambda_s = quant * var(Y_std) / nu

hyperpar = c()
hyperpar['sigmasq_mu'] = (0.5/(2*sqrt(M)))^2
hyperpar['lambda_s'] = lambda_s
hyperpar['nu'] = nu
hyperpar['lambda_k'] = 4
hyperpar['M'] = M
hyperpar['k_max'] = k_max

# MCMC parameters
# number of posterior samples = (MCMC - BURNIN) / THIN
MCMC = 10000    # MCMC iterations !! Down from 30000 originally !!
BURNIN = 5000  # burnin period length !! Down from 15000 originally !!
THIN = 5        # thinning intervals

```

Here is an implementation written using the BASTION package (based on the one from the paper)

```{r BASTION-implement}
BASTIONfit = function(Y, graph, init_vals, hyperpars, MCMC, BURNIN, THIN, seed = NULL) {
  if(!is.null(seed)) {
    set.seed(seed)
  }
    n = vcount(graph)
  # hyper-parameter
  M = hyperpars['M']  # number of trees
  sigmasq_mu = hyperpars['sigmasq_mu']
  lambda_s = hyperpars['lambda_s']
  nu = hyperpars['nu']
  lambda_k = hyperpars['lambda_k']
  k_max = hyperpars['k_max']
  hyper = c(sigmasq_mu, lambda_s, nu, lambda_k)
  
  # initial values
  mstgraph_lst = init_vals[['trees']]
  mu = init_vals[['mu']]
  cluster = init_vals[['cluster']]  # n*M matrix
  sigmasq_y = init_vals[['sigmasq_y']]
  k = as.numeric(apply(cluster, 2, max))  # number of clusters
  csize = list() # cluster size
  g = matrix(0, nrow = n, ncol = M)  # n*M matrix of fitted mu's
  for(m in 1:M) {
    cluster_m = cluster[, m]
    g[, m] = mu[[m]][cluster_m]
    csize[[m]] = Rfast::Table(cluster_m)
    mstgraph_m = mstgraph_lst[[m]]
  }
  
  # whether an edge in graph is within a cluster or between two clusters
  # n*M matrix
  
  ################# MCMC ####################
  
  ## MCMC results
  mu_out = list()
  sigmasq_y_out = numeric((MCMC-BURNIN)/THIN)
  cluster_out = array(0, dim = c((MCMC-BURNIN)/THIN, n, M))
  tree_out = list()
  log_post_out = numeric((MCMC-BURNIN)/THIN)
  
  ## MCMC iteration
  for(iter in 1:MCMC) {
    for(m in 1:M) {
      k_m = k[m]
      mstgraph_m = mstgraph_lst[[m]]
      cluster_m = cluster[, m]
      csize_m = csize[[m]]
      
      e_m = Y - Rfast::rowsums(g[, -m])
      
      if(k_m == 1) {
        rb = 0.9; 
        rd = 0; 
        rc = 0; 
        rhy = 0.1
      } else if(k_m == min(k_max, n)) {
        rb = 0;
        rd = 0.6;
        rc = 0.3;
        rhy = 0.1
      } else {
        rb = 0.3;
        rd = 0.3;
        rc = 0.3;
        rhy = 0.1
      }
      move = sample(4, 1, prob = c(rb, rd, rc, rhy))
      # current_comp = components(mstgraph_m)
      
      if(move == 1) { ## Birth move
        # split an existing cluster
        clust_split = sample.int(k_m, 1, prob = csize_m - 1)
        proposed_output = graphBirth(mstgraph_m, cluster_m, clust_split)
        vid_new = proposed_output$new_clust_ids
        vid_old = proposed_output$old_clust_ids
        
        # compute log-prior ratio
        log_A = log(lambda_k) - log(k_m + 1)
        # compute log-proposal ratio
        if(k_m == min(k_max, n)-1) {
          rd_new = 0.6
        } else {
          rd_new = 0.3
        }
        log_P = log(rd_new) - log(rb)
        # compute log-likelihood ratio
        ##
        sigma_ratio = sigmasq_y / sigmasq_mu
        csize_old = length(vid_old)
        csize_new = length(vid_new)
        sum_e_old = sum(e_m[vid_old])
        sum_e_new = sum(e_m[vid_new])
        logdetdiff = -0.5 * (log(csize_old + sigma_ratio) + 
                             log(csize_new + sigma_ratio) - 
                             log(csize_old + csize_new + sigma_ratio) - 
                             log(sigma_ratio))
        quaddiff = 0.5 * ( sum_e_old^2 / (csize_old + sigma_ratio) + 
                           sum_e_new^2 / (csize_new + sigma_ratio) -
                           (sum_e_old + sum_e_new)^2 / (csize_old + csize_new + sigma_ratio)
                          ) / sigmasq_y
        ##
        log_L = logdetdiff + quaddiff
        
        #acceptance probability
        acc_prob = min(0, log_A + log_P + log_L)
        acc_prob = exp(acc_prob)
        if(runif(1) < acc_prob){
          # accept
          mstgraph_lst[[m]] = proposed_output$graph
          csize[[m]] = components(mstgraph_m)$csize
          cluster[, m] = proposed_output$membership
          k[m] = k[m] + 1
        }
      }
      
      if(move == 2) { ## Death move
        # merge two existing clusters (c1, c2) -> c2
        proposed_output = graphDeath(mstgraph_m, cluster_m, graph)
        vid_old = proposed_output$old_clust_ids
        vid_new = proposed_output$new_clust_ids
        
        # compute log-prior ratio
        log_A = -log(lambda_k) + log(k_m)
        # # compute log-proposal ratio
        if(k_m == 2) {
          rb_new = 0.9
        }else {
          rb_new = 0.3
        }
        log_P = -(log(rd) - log(rb_new))
        # compute log-likelihood ratio
        sigma_ratio = sigmasq_y / sigmasq_mu
        csize_old = length(vid_old)
        csize_new = length(vid_new)
        sum_e_old = sum(e_m[vid_old])
        sum_e_new = sum(e_m[vid_new])
        logdetdiff = -0.5 * (log(csize_new+sigma_ratio) - 
                             log(csize_old+sigma_ratio) - 
                             log(csize_new-csize_old+sigma_ratio) + 
                             log(sigma_ratio))
        quaddiff = 0.5 * (  sum_e_new^2/(csize_new+sigma_ratio) - 
                            sum_e_old^2/(csize_old+sigma_ratio) - 
                            (sum_e_new-sum_e_old)^2/(csize_new-csize_old+sigma_ratio)
                          ) / sigmasq_y
        log_L = logdetdiff + quaddiff
        
        # acceptance probability
        acc_prob = min(0, log_A + log_P + log_L)
        acc_prob = exp(acc_prob)
        if(runif(1) < acc_prob){
          # accept
          mstgraph_m = proposed_output$graph
          csize[[m]] = components(mstgraph_m)$csize
          cluster[, m] = proposed_output$membership
          k[m] = k[m] - 1
        }
      }
      
      if(move == 3) { ## change move
        # # first perform death move: (c1, c2) -> c2
        # merge_res = mergeCluster(mstgraph_m, eid_btw_mst_m, subgraphs_m, csize_m, 
        #                          cluster_m, inc_mat, change = T)
        # # then perform birth move
        # split_res = splitCluster(mstgraph_m, k_m-1, merge_res$subgraphs, merge_res$csize)
        
        proposed_output = graphChange(mstgraph_m, cluster_m, graph)
        
        # compute log-likelihood ratio
        # First do it for the death move
        sigma_ratio = sigmasq_y / sigmasq_mu
        csize_old_d = length(vid_old_d)
        csize_new_d = length(vid_new_d)
        sum_e_old_d = sum(e_m[vid_old_d])
        sum_e_new_d = sum(e_m[vid_new_d])
        logdetdiff_d = -0.5 * (log(csize_new_d+sigma_ratio) - 
                             log(csize_old_d+sigma_ratio) - 
                             log(csize_new_d-csize_old_d+sigma_ratio) + 
                             log(sigma_ratio))
        quaddiff_d = 0.5 * (  sum_e_new_d^2/(csize_new_d+sigma_ratio) - 
                            sum_e_old_d^2/(csize_old_d+sigma_ratio) - 
                            (sum_e_new_d-sum_e_old_d)^2/(csize_new_d-csize_old_d+sigma_ratio)
                          ) / sigmasq_y
        log_L_death = logdetdiff_d + quaddiff_d
        # Now do it for the birth move
        csize_old_b = length(vid_old_b)
        csize_new_b = length(vid_new_b)
        sum_e_old_b = sum(e_m[vid_old_b])
        sum_e_new_b = sum(e_m[vid_new_b])
        logdetdiff_b = -0.5 * (log(csize_old_b + sigma_ratio) + 
                             log(csize_new_b + sigma_ratio) - 
                             log(csize_old_b + csize_new_b + sigma_ratio) - 
                             log(sigma_ratio))
        quaddiff_b = 0.5 * ( sum_e_old_b^2 / (csize_old_b + sigma_ratio) + 
                           sum_e_new_b^2 / (csize_new_b + sigma_ratio) -
                           (sum_e_old_b + sum_e_new_b)^2 / (csize_old_b + csize_new_b + sigma_ratio)
                          ) / sigmasq_y
        log_L_birth = logdetdiff_b + quaddiff_b
        # Add them together
        log_L = log_L_birth + log_L_death
        
        # acceptance probability
        acc_prob = min(0, log_L)
        acc_prob = exp(acc_prob)
        if(runif(1) < acc_prob){
          # accept
          mstgraph_lst[[m]] = proposed_output$graph
          csize[[m]] = components(mstgraph_m)$csize
          cluster[, m] = proposed_output$membership
        }
      }
      
      if(move == 4) {
        # update MST
        proposed_output = graphHyper(graph, cluster_m)
        mstgraph_lst[[m]] = proposed_output$graph
        cluster[, m] = proposed_output$membership
      }
      
      # update mu_m
      k_m = k[m]
      cluster_m = cluster[, m]
      csize_m = csize[[m]]
      Qinv_diag = 1 / (csize_m/sigmasq_y + 1/sigmasq_mu)
      b = Qinv_diag * Rfast::group.sum(e_m, cluster_m) / sigmasq_y
      mu[[m]] = rnorm(k_m, b, sqrt(Qinv_diag))
      
      g[, m] = mu[[m]][cluster_m]
    }
    
    # update sigmasq_y
    Y_hat = g[, M] + Y - e_m
    rate = 0.5*(nu*lambda_s + sum((Y - Y_hat)^2))
    sigmasq_y = 1/rgamma(1, shape = (n+nu)/2, rate = rate)
    
    
    ## save result
    if(iter > BURNIN & (iter - BURNIN) %% THIN == 0) {
      mu_out[[(iter-BURNIN)/THIN]] = mu
      sigmasq_y_out[(iter-BURNIN)/THIN] = sigmasq_y
      #tree_out[[(iter-BURNIN)/THIN]] = mstgraph_lst
      cluster_out[(iter-BURNIN)/THIN, , ] = cluster
      
      log_post_out[(iter-BURNIN)/THIN] = evalLogPost(mu, g, sigmasq_y, k, Y, hyper)
    }
    
    if(iter %% 100 == 0)
      cat('Iteration', iter, 'done\n')
  }
  
  mode(cluster_out) = 'integer'  # to save memory
  return(list('mu_out' = mu_out,
              'sigmasq_y_out' = sigmasq_y_out,
              'cluster_out' = cluster_out, 'log_post_out' = log_post_out))
}
```

Let's take a look at the speed difference:

```{r aral-speed}
# aral_speed = microbenchmark(
#   fitBAST(Y_std, graph0, init_val, hyperpar, MCMC, BURNIN, THIN, seed = 12345),
#   BASTIONfit(Y_std, graph0, init_val, hyperpar, MCMC, BURNIN, THIN, seed = 12345),
#   times = 10
# )
```


```{r aral-speed-xtable, include=FALSE}

```







